Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
syntax and semantics

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  First, the
  runtime resource cost of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  Third, an automatic amortized resource
  analysis for C programs that can compute derivations in the
  quantitative Hoare logic is developed and implemented.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.

@@

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
1043$).
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature which were obtained using
methods based on abstract interpretation.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations on an abstract domain to obtain loop invariants.
The mechanism we designed is able to leverage local assertions such as
$x < y$ that we collect along the branching points of the program to
obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
integer variables, and mutually-recursive functions.

@@

\section{Syntax}

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

The expressions are left abstract in this presentation.  To
present both the semantics and the program logic, we simply
need an evaluation function that maps any $e \in E$ to
a value $\llbracket e \rrbracket_\state$ using $\state$ as current
state.

\section{Semantics}

The operational semantics uses a mutable state $\state$
composed of two maps from variable names to integers. The
first map $\theta$ represents the local variable of a
function and the second map $\gamma$ gives values to global
variables of the program.

For simplicity, we assume that local variables and global
variables are always disjoint sets, this simplifies both
the semantics and the description of our analysis.  The
mutable state $\state$ can be accessed and modified, the
access is written functionally and is defined by
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
We note $\state[x \mapsto v]$ the state that results from
updating $\state$ to make $x$ map to $v$.  Similarly to
the state access, the formal definition of this operation
is a simple case disjunction.  Thanks to the assumption we
made above we can be sure that the different cases never
overlap.

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We achieve this
independence by parameterizing by a resource
\emph{metric}, it is a tuple of nine rational numbers and
one map from expressions to rational numbers
$$
(
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e
).
$$
Each of these rational numbers indicates the amount of resource
consumed by one operation.  If one of these numbers is negative,
it means that resources are released by the program.

For technical reasons in our soundness proof and to be in line
with CompCert's Clight, our language's semantics is defined
in a small-step fashion.  The \pref{fig:opsem} contains the full
list of reduction rules.  These rules provide a rewrite
system for program configurations of the form
$(\state, S, \cont, \cost)$, where $\state$ is the program state,
$S$ is the statement being executed, $\cont$ is a continuation
that describes what remains to be done once the execution of $S$
is over, and $\cost$ is the non-negative quantity of resources
available for further execution.  All rules that can decrease
$\cost$ have the implicit side condition that the resource
quantity available after the step is non-negative.  Using this
resource counter we can define the resource consumption of a
program as the minimal resource counter $\cost_*$ that lets the
program run to termination without errors.

The small-step semantics uses continuations to represent the
context of execution of a given statement.
A continuation can be a sequenced computation, it can also result
from entering a loop or a function, or it can simply be the
empty continuation (used to start a program).
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop \cont, \cost)
\smallstep
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop \cont, \cost)
\smallstep
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \delta f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep
(\state, S, \cont, \cost {-} M_q)
}
%

>>>>

  This paper describes a novel quantitative resource analysis
  framework that tackles these challenges for C programs.  First, the
  runtime resource cost of programs is formalized with a parametric
  cost semantics for CompCert Clight.  Second, a quantitative Hoare
  logic that can be used for interactive bound development is
  implement and proved sound with respect to the cost semantics in the
  Coq Proof Assistant.  Third, an automatic amortized resource
  analysis for C programs that computes derivations in the
  quantitative Hoare logic is developed and implemented.  Both, the
  quantitative logic and the automatic amortized analysis are
  naturally compositional and can be combined to semi-automatically
  derive global resource bounds.

@@

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ (for example $B(100,200,43) =
1043$).  The use of interval sizes in potential functions is an
important novelty.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature which were obtained using
methods based on abstract interpretation.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations on an abstract domain to obtain loop invariants.
The mechanism we designed is able to leverage local assertions such as
$x < y$ that we collect along the branching points of the program to
obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
integer variables, and mutually-recursive functions.

@@

\section{Syntax and Semantics}

We implemented our cost semantics and the quantitative Hoare logic in
Coq for \emph{CompCert Clight}.  Clight is the most abstract
intermediate language used by CompCert.  Mainly, it is a subset of C
in which loops can only be exited with a \code{break} statement and
expressions are free of side effects.

\paragraph{Syntax}

In this article, we describe our system for a subset of Clight to
discuss general ideas instead of technical details.  This subset is
given by the following grammar.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~x
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\mid \code{tick}(n)
\end{align*}
%
Expressions are $E$ are left abstract in our presentation.  For our
analysis framework, it is only important that they are side-effect
free.
%
The most notable difference to full Clight is that we can only assign
to variables and thus do not consider operations that update the heap.
Moreover, function arguments and return values are assumed to be
variables.  In the implementation we can deal with heap updates and
general function calls and returns.  However, we have not implemented
function pointers, goto statements, continue statements, and switch
statements.

We include the built-in primitive $\code{assert}~E$ that terminates
the program if the argument $E$ evaluates to false and has no effect
otherwise.  This is useful to express assumptions on the inputs of a
program for the automatic analysis.  We also add the built-in function
$\code{tick}(n)$ that can be called with a constant integer $n$ as a
flexible way to model resource consumption or release (if $n$ is
negative).

\paragraph{Semantics}

CompCert Clight's operational semantics is based on small-step
transitions and continuations.  Expressions---which do not have side
effects---are evaluated in a big-step fashion.  Here, we describe a
simplified version of Clight's semantics that is sufficient for the
subset that we are describing.

A program state $\state = (\theta, \gamma)$ composed of two maps from
variable names to integers. The first map $\theta : \text{Locals} \to
\Z$ assigns integers to local variables of a function and the second
map $\gamma : \text{Globals} \to \Z$ gives values to global variables
of the program.
%
The evaluation function \evalE{\cdot}{} that maps an expression $e \in
E$ to a value $\evalE{e}{\state} \in \Z$ in program state $\state$.

For simplicity, we assume that local variables and global
variables are always disjoint sets. % This simplifies both
% the semantics and the description of our analysis.  
The mutable state $\state$ can be accessed and modified, the access is
written functionally and is defined by
$$
(\theta, \gamma)(x) =
\left\{
\begin{array}{lr}
\theta(x) & \mbox{ if } x \in \dom \theta \\
\gamma(x) & \mbox{ if } x \in \dom \gamma
\end{array}
\right..
$$
For a program state $\state$, we write $\state[x \mapsto v]$ for the
program state that maps $x$ to $v$ and behaves as $\state$ for all
other variables, regardless whether $x \in \dom \theta$ or $x \in \dom
\gamma$.  Similarly, we just write $\state(x)$ to obtain the value of
$x$ in program state $\state$.

The small-step semantics is standard, except that it tracks the
resource consumption of a program.  The semantics is oblivious to the
resource of interest for the user of our system.  We achieve this
independence by parameterizing evaluations with a resource
\emph{metric} $M$; a tuple of nine rational numbers and one map $M_e :
\text{Expr} \to \Q$ from expressions to rational numbers.
$$
M = (
M_a,
M_b,
M_r,
M_u,
M_f,
M_l,
M_c^1,
M_c^2,
M_q,
M_e(E)
)
$$
Each of these rational numbers indicates the amount of resource
consumed by a corresponding operation.  If one of these numbers is
negative, it means that resources are released by the program.  The
metric that we use in the implementation can also depend on other
information that is statically available such as the name of the called
function or the number of arguments.

\pref{fig:opsem} contains the reduction rules of the semantics.
The rules define a rewrite system for program configurations of the
form $(\state, S, \cont, \cost)$, where $\state$ is the program state,
$S$ is the statement being executed, $\cont$ is a continuation that
describes what remains to be done after the execution of $S$, and
$\cost$ is the non-negative number of resources available for further
execution.  All rules that can decrease $\cost$ have the implicit side
condition that the resource quantity available after the step is
non-negative.

A continuation $\cont$ represents the context of the execution of a
given statement.  A continuation can be the empty continuation
$\Kstop$ (used to start a program), a sequenced computation $\Kseq S
\cont$, a loop continuation $\Kloop S \cont$, or the continuation
$\Kcall r \theta \cont$ of a function call.
$$
\cont := \Kstop
\mid \Kseq S \cont
\mid \Kloop S \cont
\mid \Kcall r \theta \cont
$$
%
During the execution we assume a fixed function context $\fenv$ that
maps function names to a list of variables and the statement that
defines the function body.  It is used in the rule \textsc{S:Call}.

The intuitive meaning of an evaluation judgement $(\state, S, \cont,
\cost) \smallstep{M}^* (\state', S', \cont', \cost')$ is the following.
If the statement $S$ is executed in program state $\state$, with
continuation $\cont$, and with $\cost$ resources available
then---after a finite number of steps---the evaluation will reach the
new machine state $(\state', S', \cont')$ and there are $\cost'$
resources available.  The resource consumption of the evaluation up to
this point is then $\cost - \cost'$.  If this difference is negative
then resource became available during the execution.

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^* (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^* (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^* (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.  Then
 $(\state, S, \cont, d) \smallstep{M}^* (\state'', \code{skip}, \Kstop,
 d')$ for some $d$, $d'$ and $\state''$.
\end{lemma}

\begin{figure}[t!]
\begin{mathpar}
%
\Rule{S:Assert}
{ \code{istrue}~\sem{e}_\state }
{ (\state, \code{assert}~e, \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_a)
}
%
\and
\Rule{S:BrkSeq}
{}
{ (\state, \code{break}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{break}, \cont, \cost)
}
%
\and
\Rule{S:BrkLoop}
{}
{ (\state, \code{break}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{skip}, \cont, \cost {-} M_b)
}
%
\and
\Rule{S:RetSeq}
{}
{ (\state, \code{return}~x, \Kseq S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
\and
%
\and
\Rule{S:RetLoop}
{}
{ (\state, \code{return}~x, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{return}~x, \cont, \cost)
}
%
\and
\Rule{S:RetCall}
{ \state = (\_, \gamma) \\
  \state' = (\theta, \gamma)[r \mapsto \state(x)]
}
{ (\state, \code{return}~x, \Kcall r \theta \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_r)
}
%
\and
\Rule{S:Update}
{ \state' = \state[x \mapsto \sem{e}_\state] }
{ (\state, x \gets e, \cont, \cost)
\smallstep{M}
(\state', \code{skip}, \cont, \cost {-} M_u {-} M_e(e))
}
%
\and
\Rule{S:Call}
{ \fenv f = (\vec x, S_f) \\
  \state = (\theta, \gamma) \\
  \state' = (\vec x \mapsto \state(\vec y), \gamma) \\
}
{ (\state, r \gets f(\vec y), \cont, \cost)
\smallstep{M}
(\state', S_f, \Kcall r \theta \cont, \cost {-} M_f)
}
%
\and
\Rule{S:Loop}
{}
{ (\state, \code{loop}~S, \cont, \cost)
\smallstep{M}
(\state, S, \Kloop S \cont, \cost)
}
%
\and
\Rule{S:SkipLoop}
{}
{ (\state, \code{skip}, \Kloop S \cont, \cost)
\smallstep{M}
(\state, \code{loop}~S, \cont, \cost {-} M_l)
}
%
\and
\Rule{S:IfTrue}
{ \code{istrue}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \cont, \cost {-} M_c^1)
}
%
\and
\Rule{S:IfFalse}
{ \code{isfalse}~\sem{e}_\sigma }
{ (\state, \code{if}(e)~S_1~\code{else}~S_2, \cont, \cost)
\smallstep{M}
(\state, S_2, \cont, \cost {-} M_c^2)
}
%
\and
\Rule{S:Seq}
{ }
{ (\state, S_1; S_2, \cont, \cost)
\smallstep{M}
(\state, S_1, \Kseq {S_2} \cont, \cost)
}
%
\and
\Rule{S:SkipSeq}
{}
{ (\state, \code{skip}, \Kseq S \cont, \cost)
\smallstep{M}
(\state, S, \cont, \cost {-} M_q)
}
%
